---
title: "Get IUCN red list data with redlist R package"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r echo=FALSE,include=FALSE}
library(redlist)
library(kableExtra)
library(dplyr)
```

# What is IUCN red list ?
The [IUCN Red List of Threatened](https://www.iucnredlist.org/) Species is a comprehensive global inventory that assesses the conservation status of plant, animal, and fungi species. Managed by the International Union for Conservation of Nature ([IUCN](https://iucn.org/)), it evaluates species based on factors like population size, rate of decline, and geographic range to classify them into categories such as Least Concern, Vulnerable, Endangered, and Critically Endangered. The Red List serves as a critical tool for conservation planning, policy-making, and raising awareness about the risk of extinction facing species worldwide. In this article, I demonstrate how to access Red List data using the redlist R package.

## Install redlist R package
The package can be installed from github with *pak* package manager. The *pak* package in R offers a faster, more reliable, and efficient way to install packages compared to traditional methods. It handles dependencies automatically, supports parallel downloads. Alternately is {devtool} or *remotes*. I prefered to use *pak*.

```{r eval=FALSE}
# Install pak if not available
if (!requireNamespace("pak", quietly = TRUE)) {
  install.packages("pak")
}

# Then install redlist
pak::pkg_install("stangandaho/redlist")
# Load the package
library(redlist)
```

## Set up an API key
If you're using this package for the first time, you'll likely need an IUCN Red List API key. You can check whether it's set by running `rl_check_api()`. If this throws an error like '*! Any redlist API available... *', you'll need to set an API key before using any of the package functions. Just follow these two simple steps:  
1. Click [here](https://api.iucnredlist.org/users/edit) to create an account if you don't already have one. Once logged in, you can generate your API key.  
2. Copy your API key and set it using the `rl_set_api()` function, like this `rl_set_api("2GoWiThmYrEDlitApiThatWorkS4me")`.  
You can then run `rl_check_api()` again to confirm that your API key is set successfully.  

## Get data
The IUCN provides extensive data through the Red List across various categories, including Assessments, Biogeographical Realms, Comprehensive Groups, Conservation Actions, Habitats, Population Trends, Red List Categories, Taxonomic Levels, and more. A complete overview, along with the corresponding access functions, can be found [here](https://stangandaho.github.io/redlist/reference/index.html#assessment). In this article, we will focus specifically on data related to threatened species that have been assessed in [Benin](https://fr.wikipedia.org/wiki/B%C3%A9nin).

```{r eval=FALSE}
benin_redlist <- rl_countries(code = "BJ")
head(benin_redlist)
```

```{r include=FALSE}
benin_redlist <- rl_countries(code = "BJ")
head(benin_redlist)
```

```{r echo=FALSE}
head(benin_redlist) %>% 
  kableExtra::kable("html") %>% 
  kableExtra::kable_styling("striped") %>% 
  kableExtra::scroll_box(width = "100%")

```
  
  
By default, the function retrieves the first page of results, which includes the first 100 records across various years. You can specify one or more pages, as well as specific publication years, to filter the data accordingly. If you want to retrieve all available data without restricting by page, you can set the page argument to `NA` or `NULL`.

```{r eval=FALSE}
# Get data from five first pages
benin_redlist <- rl_countries(code = "BJ", page = 1:5)
# Get data from two first pages specifically for 2023
benin_redlist <- rl_countries(code = "BJ", page = 1:5, year_published = 2023)
# Get all data on Benin
benin_redlist <- rl_countries(code = "BJ", page = NA)
```

As displayed, each output from *redlist* functions is returned as a tibble, allowing us to easily filter observations for analytical purposes. In the article on [data visulisation](https://stangandaho.github.io/redlist/articles/data_visualisation.html), I’ll present a few examples of plots to illustrate how the data can be explored visually.
